---
title: "Enrichment plots"
author:
  - "Jelmer Poelstra"
date: today
format:
  html:
    self-contained: true
    cache: false
    code-fold: true
    number-sections: true
    number-depth: 2
    toc: true
    fig-dpi: 300
    df-print: paged
    grid:
      body-width: 800px
knitr:
  opts_chunk:
    out.width: "100%"
    message: FALSE
execute: 
  warning: false
editor_options: 
  chunk_output_type: console
---

```{r}
# SET-UP -----------------------------------------------------------------------
# Load packages
library(DESeq2)
library(tidyverse)
library(patchwork)
library(here)

# Load script with functions
source(here("mcic-scripts/rnaseq/rfuns/DE_funs.R"))
source(here("mcic-scripts/rnaseq/rfuns/enrich_funs.R"))
source(here("mcic-scripts/misc/qmd_funs.R"))

# Settings
theme_set(theme_bw(base_size = 13))
theme_update(panel.grid.minor = element_blank())

# Define input files
GO_file <- here("results/GO/GO_res.tsv")
KEGG_file <- here("results/KEGG/KEGG_res.tsv")

# Define output files
outdir <- here("results/figs-ms")
dir.create(outdir, recursive = TRUE, showWarnings = FALSE)
GOplot_file <- file.path(outdir, "GO-plot.png")
KEGGplot_file <- file.path(outdir, "KEGG-plot.png")
```

```{r}
# Read inputs ------

# GO
GO <- read_tsv(GO_file, show_col_types = FALSE) |>
  # Use "downregulated" instead of "down", etc, so the facet labels will be clearer:
  mutate(
    DE_direction = ifelse(
      DE_direction == "down",
      "downregulated",
      "upregulated"
    )
  )

# KEGG
KEGG <- read_tsv(KEGG_file, show_col_types = FALSE) |>
  mutate(
    DE_direction = ifelse(
      DE_direction == "down",
      "downregulated",
      "upregulated"
    )
  )
```

## GO

### Number of significant terms

```{r}
GO |> count(DE_direction)
```

### Table

- `fold_enrich`: Fold enrichment (overrepresentation) of the GO term among DE genes
- `n_focal_in_cat`: Number of DE genes in the GO term
- `n_cat`: Number of genes in the GO term
- `median_lfc`: Median Log2-fold change among DE genes in the GO term

```{r}
GO |>
  select(
    DE_direction,
    term,
    description,
    padj,
    fold_enrich,
    n_focal_in_cat,
    n_cat,
    median_lfc
  ) |>
  make_dt()
```

### 15 most significant BP terms for each DE direction

```{r, fig.height=8}
# Subset GO terms
GO_sel <- GO |>
  # Only take Biological Process in this example:
  filter(ontology == "BP") |>
  # Only take the 15 most significant terms for each DE direction
  slice_min(padj, n = 15, with_ties = FALSE, by = "DE_direction") |>
  # Sorting here will be respected in the plot - sort by fold enrichment, the axis variable:
  arrange(fold_enrich)

# Plot
GO_sel |>
  cdotplot(
    x_var = "fold_enrich",
    fill_var = "padj_log",
    label_var = "n_focal_in_cat",
    facet_var1 = "DE_direction",
    facet_to_columns = FALSE,
    add_term_id = FALSE,
    label_chars = 60,
    ylab_size = 10
  ) +
  scico::scale_color_scico(
    palette = "batlow",
    name = expression("-Log"[10] * " P")
  )

#ggsave(GOplot_file, width = 9, height = 7, dpi = 600)
```

### 15 smallest (most specific) BP terms for each DE direction

```{r, fig.height=8}
# Subset GO terms
GO_sel <- GO |>
  # Only take Biological Process in this example:
  filter(ontology == "BP") |>
  # Only take the 15 'smallest terms' (containing fewest genes => more specific) per DE direction:
  slice_min(n_cat, n = 15, with_ties = FALSE, by = "DE_direction") |>
  # Sorting here will be respected in the plot - sort by fold enrichment, the axis variable:
  arrange(fold_enrich)

# Plot
GO_sel |>
  cdotplot(
    x_var = "fold_enrich",
    fill_var = "padj_log",
    label_var = "n_focal_in_cat",
    facet_var1 = "DE_direction",
    facet_to_columns = FALSE,
    add_term_id = FALSE,
    label_chars = 60,
    ylab_size = 10
  ) +
  scico::scale_color_scico(
    palette = "batlow",
    name = expression("-Log"[10] * " P")
  )

#ggsave(GOplot_file, width = 9, height = 7, dpi = 600)
```

### 5 most significant terms for each DE direction and each ontology

```{r, fig.height=8}
# Subset GO terms
GO_sel <- GO |>
  # Only take Biological Process in this example:
  # Only take the 15 most significant terms for each DE direction
  slice_min(
    padj,
    n = 5,
    with_ties = FALSE,
    by = c("DE_direction", "ontology")
  ) |>
  # Sorting here will be respected in the plot - sort by fold enrichment, the axis variable:
  arrange(DE_direction, fold_enrich)

# Plot
GO_sel |>
  cdotplot(
    x_var = "fold_enrich",
    fill_var = "padj_log",
    label_var = "n_focal_in_cat",
    facet_var1 = "ontology",
    facet_var2 = "DE_direction",
    add_term_id = FALSE,
    label_chars = 60,
    ylab_size = 10
  ) +
  scico::scale_color_scico(
    palette = "batlow",
    name = expression("-Log"[10] * " P")
  )

#ggsave(GOplot_file, width = 9, height = 7, dpi = 600)
```

## KEGG

```{r}
# Plot
KEGG |>
  arrange(DE_direction, fold_enrich) |>
  cdotplot(
    x_var = "fold_enrich",
    fill_var = "padj_log",
    label_var = "n_focal_in_cat",
    facet_var1 = "DE_direction",
    facet_to_columns = FALSE,
    add_term_id = FALSE,
    label_chars = 60,
    ylab_size = 10
  ) +
  scico::scale_color_scico(
    palette = "batlow",
    name = expression("-Log"[10] * " P")
  )
```

## Session info

```{r}
sessionInfo()
```
